Multiproxy analysis exploring patterns of diet and disease in dental calculus and skeletal remains from a 19th century Dutch population

### 2023-06-01 analysis/paper/_discussion.qmd
fix broken citekeys

<!-- compounds detected with potential sources, and those that were surprisingly missing -->
<!-- nicotine and cotinine: tobacco, obviously...-->
Nicotine and its principal/main metabolite, cotinine, were strongly positively correlated,
both in concentration and presence/absence in individuals (@tbl-pearson and @fig-polycorr).
The detection of nicotine and cotinine is not surprising, as pipe-smoking in the
Beemsterpolder is well-documented in the literature
[@aten400JaarBeemster2012; @boumanBegravenis2017], and visible on the skeletal
remains as pipe notches [@lemmersMiddenbeemster2013].
There is also documented medicinal use of nicotine in the Beemsterpolder,
where a tobacco-smoke enema was used for headaches, respiratory problems, colds,
and drowsiness from around 1780 to 1830 [@aten400JaarBeemster2012].
In our sample, we also detected nicotine and cotinine (replicated)
in an old adult, probable female individual.
In this particular case it is unlikely that the compounds entered the dental
calculus through pipe-smoking, as the individual had no visible pipe notches;
more likely the tobacco entered through an alternate mode of consumption,
secondhand smoke, or the aforementioned tobacco-smoke enema.

@@

<!-- salicylic acid: diet, medicine, or environmental??? -->
Salicylic acid was found in all but one individual in our sample.
It can be extracted from the bark of willow trees, *Salix alba*, and
has long been used for its
pain-relieving properties [@bruinsmaBijdragenTot1872, p. 119]. It is also present
in many plant-based foods
[@duthieNaturalSalicylates2011; @malakarNaturallyOccurring2017], including
potatoes, which were a staple of the Beemsterpolder diet [@aten400JaarBeemster2012].
The extracted quantity from our samples decreased over the three washes, followed
by a sharp increase in the final calculus extraction, which is what we would expect
to see if the salicylic acid was incorporated during life [@fig-auth-plot-batch2].
However, it has been shown that salicilyc acid is a very mobile organic acid and
the ubiquitous presence may be due to environmental contamination, which would
also explain the high quantity in the washes
[@chenCa2Dependent2001; @badriRegulationFunction2009].
Given the multiple plausible sources of this residue, it will be necessary to
explore the extent to which salicylic acid can leach
into the dental calculus from the soil, and what the rate of degradation is for
salicylic acid when trapped in dental calculus.
<!-- perhaps using an oral biofilm model ;) -->

@@

<!-- opioids: Damn... that would've been awesome -->
Despite many of our sampled individuals having lived during the height of the opium era
in the Netherlands [@machtHistoryOpium1915], none of the targeted opioids
(morphine, codeine, thebaine, papaverine, norcodeine, noscapine)
were detected. 
The absence of opioids could be a result of the people ascribing more to the
"traditional" rather than "scientific" medicine, although laudanum and
another opium containing concoction was part of the "traditional" medicine in
the Netherlands [@leuwProhibitionLegalizationDutch1994], including Middenbeemster
[@aten400JaarBeemster2012]. It was also generally
considered a drug of the upper class [@scheltemaOpiumTrade1907], and may have been
more common in urban centers.
The absence could also be attributed to postmortem degradation.
It has been shown that, while abundant in opium, morphine
degrades rapidly, while thebaine and papaverine are more resistant to various
ageing processes [@chovanecOpiumMasses2012]. The latter were also absent from
our samples.

>>>>

<!-- compounds detected with potential sources, and those that were surprisingly missing -->
<!-- nicotine and cotinine: tobacco, obviously...-->
Nicotine and its principal/main metabolite, cotinine, were strongly positively correlated,
both in concentration and presence/absence in individuals (@tbl-pearson and @fig-polycorr).
The detection of nicotine and cotinine is not surprising, as pipe-smoking in the
Beemsterpolder is well-documented in the literature
[@aten400Jaar2012; @boumanBegravenis2017], and visible on the skeletal
remains as pipe notches [@lemmersMiddenbeemster2013].
There is also documented medicinal use of nicotine in the Beemsterpolder,
where a tobacco-smoke enema was used for headaches, respiratory problems, colds,
and drowsiness from around 1780 to 1830 [@aten400Jaar2012].
In our sample, we also detected nicotine and cotinine (replicated)
in an old adult, probable female individual.
In this particular case it is unlikely that the compounds entered the dental
calculus through pipe-smoking, as the individual had no visible pipe notches;
more likely the tobacco entered through an alternate mode of consumption,
secondhand smoke, or the aforementioned tobacco-smoke enema.

@@

<!-- salicylic acid: diet, medicine, or environmental??? -->
Salicylic acid was found in all but one individual in our sample.
It can be extracted from the bark of willow trees, *Salix alba*, and
has long been used for its
pain-relieving properties [@bruinsmaBijdragenTot1872, p. 119]. It is also present
in many plant-based foods
[@duthieNaturalSalicylates2011; @malakarNaturallyOccurring2017], including
potatoes, which were a staple of the Beemsterpolder diet [@aten400Jaar2012].
The extracted quantity from our samples decreased over the three washes, followed
by a sharp increase in the final calculus extraction, which is what we would expect
to see if the salicylic acid was incorporated during life [@fig-auth-plot-batch2].
However, it has been shown that salicilyc acid is a very mobile organic acid and
the ubiquitous presence may be due to environmental contamination, which would
also explain the high quantity in the washes
[@chenCa2Dependent2001; @badriRegulationFunction2009].
Given the multiple plausible sources of this residue, it will be necessary to
explore the extent to which salicylic acid can leach
into the dental calculus from the soil, and what the rate of degradation is for
salicylic acid when trapped in dental calculus.
<!-- perhaps using an oral biofilm model ;) -->

@@

<!-- opioids: Damn... that would've been awesome -->
Despite many of our sampled individuals having lived during the height of the opium era
in the Netherlands [@machtHistoryOpium1915], none of the targeted opioids
(morphine, codeine, thebaine, papaverine, norcodeine, noscapine)
were detected. 
The absence of opioids could be a result of the people ascribing more to the
"traditional" rather than "scientific" medicine, although laudanum and
another opium containing concoction was part of the "traditional" medicine in
the Netherlands [@leuwProhibitionLegalization1994], including Middenbeemster
[@aten400Jaar2012]. It was also generally
considered a drug of the upper class [@scheltemaOpiumTrade1907], and may have been
more common in urban centers.
The absence could also be attributed to postmortem degradation.
It has been shown that, while abundant in opium, morphine
degrades rapidly, while thebaine and papaverine are more resistant to various
ageing processes [@chovanecOpiumMasses2012]. The latter were also absent from
our samples.

### 2023-06-01 analysis/paper/_materials.qmd
fix broken citekeys

<!-- Demographics and description of MB11 sample, selection criteria. -->
The sample consists of `r nrow(demography)` individuals from Middenbeemster,
a 19th century rural Dutch site.
The village of Middenbeemster and the surrounding Beemsterpolder was established
in the beginning of the 17th century, when the Beemster lake was drained to
create more farmland, mainly for the cultivation of cole seeds (de Vries 1978).
In 1615, a decision was made to build a church, and
construction started in 1618 (Hakvoort 2013). The excavated cemetery is associated
with the Keyserkerk church, where the inhabitants of the Middenbeemster village
and the surrounding Beemsterpolder were buried between AD 1615 and 1866
[@lemmersMiddenbeemster2013]. Archival documents are available for those buried
between AD 1829 and 1866, when the majority of individuals were interred
[@palmerActivityReconstruction2016]. The main occupation of the inhabitants
was dairy farming, consisting largely of manual labour prior to the industrial
revolution [@aten400JaarBeemster2012; @palmerActivityReconstruction2016].

>>>>

<!-- Demographics and description of MB11 sample, selection criteria. -->
The sample consists of `r nrow(demography)` individuals from Middenbeemster,
a 19th century rural Dutch site.
The village of Middenbeemster and the surrounding Beemsterpolder was established
in the beginning of the 17th century, when the Beemster lake was drained to
create more farmland, mainly for the cultivation of cole seeds (de Vries 1978).
In 1615, a decision was made to build a church, and
construction started in 1618 (Hakvoort 2013). The excavated cemetery is associated
with the Keyserkerk church, where the inhabitants of the Middenbeemster village
and the surrounding Beemsterpolder were buried between AD 1615 and 1866
[@lemmersMiddenbeemster2013]. Archival documents are available for those buried
between AD 1829 and 1866, when the majority of individuals were interred
[@palmerActivityReconstruction2016]. The main occupation of the inhabitants
was dairy farming, consisting largely of manual labour prior to the industrial
revolution [@aten400Jaar2012; @palmerActivityReconstruction2016].
